### 2018-02-16 manuscript.Rnw
typos, figures and citations (a bit of everything)

% Title must be 250 characters or less.
\begin{flushleft}
{\Large
\textbf\newline{Rethomics: an R framework to analyse high-throughput behavioural data} 
}
\newline
% Insert author names, affiliations and corresponding author email (do not include titles, positions, or degrees).
\\
Quentin Geissmann\textsuperscript{1*},
Luis Garcia Rodriguez\textsuperscript{2},
Esteban J Beckwith\textsuperscript{1},
Giorgio F Gilestro\textsuperscript{1*}
\\
\bigskip
\textbf{1} Department of Life Sciences, Imperial College London, London, United Kingdom
\\
\textbf{2} Affiliation Dept/Program/Center, Institution Name, City, State, Country %TODO @luis
\\
\bigskip

@@

The availability of large amounts of data is very exciting as it paves the way for in-depth quantitative analyses.
The multiplicity of model organisms, hypotheses and paradigms makes the existence of diverse range of recording tools important.
However, when it comes to the subsequent processing of the results, there is no unified, programmatic, framework that could be used as a set of building blocks in a pipeline.
Instead, tools tend to consist of graphical interfaces that tend to import data from a single platform and provide somewhat rigid functionalities.
There are, at least, three issues with this approach.
First of all, state-of-the-art analysis and visualisation require a level of reproducibility, flexibility and scalability that only a programming interface can provide.
Secondly, it favours replicated work as developers need to create their independent solution to similar problems.
Lastly, it links analysis and visualisation to the target acquisition tool, which makes it very difficult to share cross-tool utilities and concepts.

Thankfully, behavioural data is conceptually largely agnostic of the acquisition platform and paradigm. 	
Typically, the behaviour of each individual is described by a long time series (possibly multivariate and irregular).
Importantly, individuals are labelled with arbitrary metadata defined by the experimenter (\emph{e.g.} sex, treatment and genotype).
Efficiently combining and manipulating metadata and data of hundreds of individuals, each recorded for days or weeks, is not trivial.

In the article herein, we describe \texttt{rethomics}, a framework that unifies analysis of behavioural dataset in an efficient and flexible manner. We implemented it in \texttt{R}\cite{r_core_team_r:_2017} since it is widely taught and adopted in the community of computational biologists with a vast number of open-source packages.
\texttt{rethomics} offers an elegant computational solution to store, manipulate and visualise  large amounts of data.
We expect it to fill the gap between behavioural biology and data sciences, thus promoting collaboration between computational and behavioural scientists.
\texttt{rethomics} comes with a extensive documentation and a set of both practical and theoretical demos and tutorials.

@@

\begin{figure}[!h]
	%\includegraphics[width=1\textwidth]{fig/fig-1.pdf}
	\caption{{\bf The \texttt{rethomics} workflow.}
		Diagram representing the interplay between, from left to right, the raw data, the \texttt{rethomics} packages (in blue) and the rest of the \texttt{R} ecosystem.}
	\label{fig:fig-1}
\end{figure}

\subsection*{Internal data structure}
We created \texttt{behavr} (Fig~\ref{fig:fig-2}A), a new data structure, based on the widely adopted \texttt{data.table} object, in order to address two challenges that are inherent to manipulating ethomics results.

@@

\begin{figure}[!h]
	%\includegraphics[width=1\textwidth]{fig/fig-2.pdf}
	\caption{{\bf \texttt{behavr} table.}
	A: Illustration of a \texttt{behavr} object, the core data structure in \texttt{rethomics}.
		The metadata holds a single row for each of the $n$ individuals. 
		Its columns, the $p$ metavariables, are one of two kinds: either required -- and defined by the acquisition platform (\emph{i.e.} used to fetch the data) -- or user-defined (\emph{i.e.} arbitrary).
		In the data, each row is a ``read'' (\emph{i.e.} information about one individual at one time-point).
		It is formed of $q$ variables and is expected to have a very large number of reads $k$ for each individual $i$.
		Data and metadata are implicitly joined on the \texttt{id} field.
		Note that the names used in this for variables and metavariable in this example are only plausible cases that will likely differ in practice. 
		B: Non exhaustive list of uses of a \texttt{behavr} table (referred as \texttt{dt}). 
		In addition to operations on data, which are inherited from \texttt{data.table},
		we provide utilities designed specifically to act on both metadata and data.  
	}
	\label{fig:fig-2}
\end{figure}

\subsection*{Data import}
Data import package translate results from a recording platform (\emph{e.g.} text files and databases) into a single \texttt{behavr} object.
Currently, we provide a package to read Drosophila Activity Monitor (DAM2)\citationneeded{} data and another one for Ethoscope data\cite{geissmann_ethoscopes:_2017}.
Although the structure of the raw results is very different, conceptually, loading data is very similar.
In all cases, the user is asked to generate a metadata table (one row per individual). 
In it, there will be both mandatory and optional columns.
The mandatory ones are the necessary and sufficient information to fetch data (\emph{e.g.} machine id, region of interest and date). 
The optional columns are user-defined arbitrary fields that translate experimental conditions (\emph{e.g.} treatment, genotype and sex).

In this respect, the metadata file is a standardised and comprehensive data frame describing an experiment.
It explicitly lists all treatments and individuals, which facilitates interspersion of treatments (indeed, without it, users are tempted to simplify their experimental design by, for instance, confounding device/location and treatment).
Furthermore, it streamlines the inclusion and analysis of further replicates in the same workflow.
Indeed, additional replicates can simply be added as new rows -- and the ID of the replicate later used, if needed, as a covariate.

>>>>

% Title must be 250 characters or less.
\begin{flushleft}
{\Large
\textbf\newline{Rethomics: an R framework to analyse high-throughput behavioural data} 
}
\newline
% Insert author names, affiliations and corresponding author email (do not include titles, positions, or degrees).
\\
Quentin Geissmann\textsuperscript{1*},
Luis G Rodriguez\textsuperscript{2},
Esteban J Beckwith\textsuperscript{1},
Giorgio F Gilestro\textsuperscript{1*}
\\
\bigskip
\textbf{1} Department of Life Sciences, Imperial College London, London, United Kingdom
\\
\textbf{2} Institute for Neuro- and Behavioral Biology, Westf{\"a}lische Wilhelms University, 48149 M{\"u}nster, Germany
\\
\bigskip

@@

The availability of large amounts of data is very exciting as it paves the way for in-depth quantitative analyses.
The multiplicity of model organisms, hypotheses and paradigms makes the existence of a diverse range of recording tools important.
However, when it comes to the subsequent processing of the results, there is no unified, programmatic, framework that could be used as a set of building blocks in a pipeline.
Instead, tools tend to consist of graphical interfaces that import data from a single platform and provide somewhat rigid functionalities.
There are, at least, three issues with this approach.
First of all, state-of-the-art analysis and visualisation require a level of reproducibility, flexibility and scalability that only a programming interface can provide.
Secondly, it favours replicated work as developers need to create their independent solution to similar problems.
Lastly, it links analysis and visualisation to the target acquisition tool, which makes it very difficult to share cross-tool utilities and concepts.

Thankfully, behavioural data is conceptually largely agnostic of the acquisition platform and paradigm. 	
Typically, the behaviour of each individual is described by a long time series (possibly multivariate and irregular).
Importantly, individuals are labelled with arbitrary metadata defined by the experimenter (\emph{e.g.} sex, treatment and genotype).
Efficiently combining and manipulating metadata and data from hundreds of individuals, each recorded for days or weeks, is not trivial.

In the article herein, we describe \texttt{rethomics}, a framework that unifies analysis of behavioural dataset in an efficient and flexible manner.
We implemented it in \texttt{R}\cite{r_core_team_r:_2017} since it is widely taught and adopted by computational biologists.
It is also complemented with a vast ecosystem of open-source packages.
\texttt{rethomics} offers an elegant computational solution to store, manipulate and visualise large amounts of data.
We expect it to fill the gap between behavioural biology and data sciences, thus promoting collaboration between computational and behavioural scientists.
\texttt{rethomics} comes with a extensive documentation and a set of both practical and theoretical demos and tutorials.

@@

\begin{figure}[!h]
%	\includegraphics[width=1\textwidth]{fig/fig-1.pdf}
	\caption{{\bf The \texttt{rethomics} workflow.}
		Diagram representing the interplay between, from left to right, the raw data, the \texttt{rethomics} packages (in blue) and the rest of the \texttt{R} ecosystem.}
	\label{fig:fig-1}
\end{figure}

\subsection*{Internal data structure}
We created \texttt{behavr} (Fig~\ref{fig:fig-2}A), a new data structure, based on the widely adopted \texttt{data.table} object, in order to address two challenges that are inherent to handling ethomics results.

@@

\begin{figure}[!h]
% 	\includegraphics[width=1\textwidth]{fig/fig-2.pdf}
	\caption{{\bf \texttt{behavr} table.}
	A: Illustration of a \texttt{behavr} object, the core data structure in \texttt{rethomics}.
		The metadata holds a single row for each of the $n$ individuals. 
		Its columns, the $p$ metavariables, are one of two kinds: either required -- and defined by the acquisition platform (\emph{i.e.} used to fetch the data) -- or user-defined (\emph{i.e.} arbitrary).
		In the data, each row is a ``read'' (\emph{i.e.} information about one individual at one time-point).
		It is formed of $q$ variables and is expected to have a very large number of reads, $k$, for each individual $i$.
		Data and metadata are implicitly joined on the \texttt{id} field.
		Note that the names used in this for variables and metavariable in this example are only plausible cases which will likely differ in practice. 
		B: Non exhaustive list of uses of a \texttt{behavr} table (referred as \texttt{dt}). 
		In addition to operations on data, which are inherited from \texttt{data.table},
		we provide utilities designed specifically to act on both metadata and data.  
		Commented examples are prefixed by \texttt{>}.
	}
	\label{fig:fig-2}
\end{figure}

\subsection*{Data import}
Data import package translate results from a recording platform (\emph{e.g.} text files and databases) into a single \texttt{behavr} object.
Currently, we provide a package to read single or multi-beam Drosophila Activity Monitor (DAM)\citationneeded{} data and another one for Ethoscope data\cite{geissmann_ethoscopes:_2017}.
Although the structure of the raw results is very different, conceptually, loading data is very similar.
In all cases, the user is asked to generate a metadata table (one row per individual). 
In it, there will be both mandatory and optional columns.
The mandatory ones are the necessary and sufficient information to fetch data (\emph{e.g.} machine id, region of interest and date). 
The optional columns are user-defined arbitrary fields that translate experimental conditions (\emph{e.g.} treatment, genotype and sex).

In this respect, the metadata file is a standardised and comprehensive data frame describing an experiment.
It explicitly lists all treatments and individuals, which facilitates interspersion of conditions (indeed, without it, users are tempted to simplify their experimental design by, for instance, confounding device/location and treatment).
Furthermore, it streamlines the inclusion and analysis of further replicates in the same workflow.
Indeed, additional replicates can simply be added as new rows -- and the ID of the replicate later used, if needed, as a covariate.
